---
title: "Advent of Code 2020"
output: 
    github_document: 
        toc: true
        toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(igraph)
library(tidygraph)
walk(list.files(here::here("R"), full.names = TRUE), source)
```

Here's my work on Advent of Code 2020. Let's see if I do more than my usual thing of getting halfway and then not finding time for the rest!

# Day 1

## Part 1

First we need the product of the two numbers in the expense report that add to 2020. 

```{r day1-part1}
day1_expense_report <- readLines(here::here("data/day1.txt")) %>% 
    as.double()

day1_expense_report %>% 
    {
        expand_grid(
            num1 = ., 
            num2 = .
        )
    } %>% 
    filter(
        # Get rid of rows where a number is paired with itself, and dedupe, by 
        # requiring that num1 < num2
        num1 < num2, 
        # Keep just the row with the right sum
        near(num1 + num2, 2020)
    ) %>% 
    pivot_longer(cols = everything()) %>% 
    pull(value) %>% 
    prod() %>% 
    {
        sprintf("Part 1 answer is %s", .)
    }
```

## Part 2

Now do the same but for the **three** numbers that add to 2020. 

```{r day1-part2}
day1_expense_report %>% 
    {
        expand_grid(
            num1 = ., 
            num2 = ., 
            num3 = .
        )
    } %>% 
    filter(
        # Get rid of rows where a number is paired with itself, and dedupe, by 
        # requiring that num1 < num2
        num1 < num2, 
        num2 < num3, 
        # Keep just the row with the right sum
        near(num1 + num2 + num3, 2020)
    ) %>% 
    pivot_longer(cols = everything()) %>% 
    pull(value) %>% 
    reduce(`*`) %>% 
    {
        sprintf("Part 2 answer is %s", .)
    }
```

# Day 2

## Part 1

Count how many of the passwords are valid. Created a function to compare a password and policy and test if it's valid. 

```{r test_password}
test_password
```

```{r day2-part1}
day2_passwords <- read_delim(
    here::here("data/day2.txt"), 
    delim = " ", 
    col_names = c("policy", "letter", "password"), 
    col_types = "ccc"
) %>% 
    # get rid of the trailing colon on the letter
    mutate(
        letter = str_remove(letter, ":")
    )
day2_passwords %>% 
    mutate(valid = pmap_lgl(., test_password)) %>% 
    filter(valid) %>% 
    {
        sprintf("Part 1 answer is %s", nrow(.))
    }
```

## Part 2

Now we need a different version of the password testing function that will comply with the Official Toboggan Corporate Policy (OTCP). 

```{r test_password_OTCP}
test_password_OTCP
```

```{r day2-part2}
day2_passwords %>% 
    mutate(valid = pmap_lgl(., test_password_OTCP)) %>% 
    filter(valid) %>% 
    {
        sprintf("Part 2 answer is %s", nrow(.))
    }
```

# Day 3

## Part 1

Need to navigate the map and count the number of trees we hit travelling **down 1, across 3** at every step. The same map repeats infinitely to the right. 

```{r day3-data}
# Save the 'raw' input: partly because its dimensions will be useful when
# creating the map, partly because I'm bound to need it again for Part 2
day3_input <- readLines(here::here("data/day3.txt"))

day3_map <- day3_input %>% 
    map(
        ~ .x %>% 
            str_split("") %>% 
            pluck(1)
    ) %>% 
    unlist() %>% 
    matrix(nrow = length(day3_input), byrow = TRUE)
```


```{r count_trees}
count_trees
```

```{r day3-part1}
sprintf("Part 1 answer is %s trees", count_trees(day3_map, 1, 3))
```

## Part 2

Now need to check a number of trajectories and then return the product of the result. 

```{r day3-part2}
down <-  c(1, 1, 1, 1, 2)
right <- c(1, 3, 5, 7, 1)

map2_int(
    down, 
    right, 
    ~ count_trees(day3_map, .x, .y)
) %>% 
    prod() %>% 
    {
        sprintf("Part 2 answer is %s", .)
    }
```
# Day 4

## Part 1

Data validation. This is a bit of a busman's holiday but off we go. 

```{r day4-setup}
# get the required fields
req <- c("byr", "iyr", "eyr", "hgt", "hcl", "ecl", "pid")
day4_input <- tibble(x = read_lines(here::here("data/day4.txt")))

day4_passports <- day4_input %>% 
    mutate(id = cumsum(x == "")) %>% 
    mutate(pairs = str_match_all(x, "(...)\\:([^ ]+)")) %>% 
    mutate(
        key = map(pairs, ~ .x[, 2]), 
        value = map(pairs, ~ .x[, 3])
    ) %>% 
    select(-c(x, pairs)) %>% 
    unnest(c(key, value))
```

Now we have the passports cleaned up it's just grouping and counting. 

```{r day4-part1}
day4_valid <- day4_passports %>% 
    group_by(id) %>% 
    filter(all(req %in% c(key)))

sprintf("Part 1 answer is %s valid passports", 
        length(unique(day4_valid$id)))
```

## Part 2

Now need to validate the values. 

```{r day4-part2}
day4_valid %>% 
    # make sure we don't have any stray spaces messing things up in the value 
    # field
    mutate(across(where(is.character), str_trim)) %>% 
    pivot_wider(names_from = "key") %>% 
    # now filter the records
    filter(
        between(as.integer(byr), 1920L, 2002L),
        between(as.integer(iyr), 2010L, 2020L), 
        between(as.integer(eyr), 2020L, 2030L),
        
        (str_detect(hgt, "cm") & between(parse_number(hgt), 150, 193)) | 
            (str_detect(hgt, "in") & between(parse_number(hgt), 59, 76)), 
        
        str_detect(hcl, "^\\#[a-f0-9]{6}$"), 
        
        ecl %in% c("amb", "blu", "brn", "gry", "grn", "hzl", "oth"), 
        
        str_detect(pid, "^\\d{9}$")
    ) %>% 
    pull(id) %>% 
    length() %>% 
    {
        sprintf("Part 2 answer is %s", .)
    }
```

# Day 5

## Part 1

The problem uses binary partitioning to find a seat on the plane. The row numbers get higher as they go from the `F`ront to the `B`ack; column numbers start low on the `L`eft and are higher on the `R`ight. 

Can do this by converting to binary. 

```{r parse_seat_spec}
parse_seat_spec
```

Now point this at today's input. 

```{r day5-part1}
day5_specs <- tibble(seat_spec = read_lines(here::here("data/day5.txt")))

day5_seats <- day5_specs %>% 
    mutate(combined = map(seat_spec, parse_seat_spec)) %>% 
    unnest_wider(combined) %>% 
    mutate(seat_id = map2_int(row, col, ~ (8L * .x) + .y))

sprintf("Part 1 answer: highest seat ID is %s", max(day5_seats$seat_id))
```

## Part 2

Need to find my seat, which is the only one missing. I know that mine won't be the lowest or highest seat ID.

```{r day5-part2}
seq(min(day5_seats$seat_id), max(day5_seats$seat_id)) %>% 
    setdiff(day5_seats$seat_id) %>% 
    {
        sprintf("Part 2 answer: my seat id is %s", .)
    }
```

# Day 6

## Part 1

Need counts for how many distinct questions were answered yes by anyone in each group. Or: the union of all the sets of answers. 

```{r day6-part1}
day6_answers <- tibble(x = read_lines(here::here("data/day6.txt"))) %>% 
    mutate(group_id = cumsum(x == "")) %>% 
    filter(x != "") %>% 
    add_count(group_id, name = "group_size") %>% 
    mutate(split_x = str_split(x, ""))

day6_answers %>% 
    group_by(group_id) %>% 
    summarise(all_yes = reduce(split_x, union), .groups = "drop") %>% 
    nrow() %>% 
    {
        sprintf("Part 1 answer is %s", .)
    }
```

## Part 2

Now need counts for how many questions were answered yes by _everyone_ in each group. Just replace `union()` with `intersect()`. 

```{r day6-part2}
day6_answers %>% 
    group_by(group_id) %>% 
    summarise(all_yes = reduce(split_x, intersect), .groups = "drop") %>% 
    nrow() %>% 
    {
        sprintf("Part 2 answer is %s", .)
    }
```

# Day 7

## Part 1

This sort of problem lends itself to some sort of graph structure. Time to learn a bit about {igraph}. 

```{r day7-part1}
day7_input <- read_lines(here::here("data/day7.txt"))

day7_tidy <- tibble(x = day7_input) %>%
    separate(x, into = c("outside", "inside"), sep = " bags contain ") %>%
    separate_rows(inside, sep = ", ") %>%
    mutate(inside = str_remove(inside, " bags?.?")) %>%
    separate(inside, into = c("weight", "inside"), extra = "merge") %>%
    mutate(weight = str_replace(weight, "no", "0") %>% parse_number())

day7_nodes <- tibble(desc = union(day7_tidy$outside, day7_tidy$inside)) %>%
    rowid_to_column()

day7_edges <- day7_tidy %>%
    left_join(day7_nodes, by = c("outside" = "desc")) %>% 
    rename(to = rowid) %>% 
    left_join(day7_nodes, by = c("inside" = "desc")) %>% 
    rename(from = rowid) %>% 
    filter(!near(weight, 0)) %>% 
    select(from, to, weight)

day7_graph <- tbl_graph(
    nodes = day7_nodes, 
    edges = day7_edges, 
    directed = TRUE
)

shiny_gold_id <- day7_nodes %>% filter(desc == "shiny gold") %>% pull(rowid)
all_simple_paths(
    day7_graph, 
    from = shiny_gold_id, 
    mode = "out"
) %>% 
    map(~ .x %>% as.integer() %>% discard(~ .x == shiny_gold_id)) %>% 
    unlist() %>% 
    unique() %>% 
    length() %>% 
    {
        sprintf("There are %s possible containers", .)
    }
```

## Part 2

Previously I had done this with recursion because I hadn't set up the graph correctly, so my answers were off, which made me give up on the graph thing. I've kept that method below for posterity. Switching to a graph makes it easier and matches the intuition of the problem. 

```{r day7-part2}
all_simple_paths(day7_graph, from = shiny_gold_id, mode = "in") %>% 
    map_dbl(
        function(x) {
            x <- unclass(x)
            tibble(to = x[seq_along(x) - 1], 
                   from = x[seq(2, length(x))]) %>% 
                left_join(day7_edges, by = c("to", "from")) %>% 
                pull(weight) %>% 
                prod()
        }
    ) %>% 
    sum() %>% 
    {
        sprintf(
            "Part 2 answer: the shiny gold bag must contain %s other bags", 
            .
        )
    }
```

Now we need to count the total number of bags that the `shiny gold` must contain. Use some recursion for this. Took me a while to get it right, forgot to add the 1 at the end. 

```{r day7-part2-old}
count_bags <- function(bag_desc) {
    d <- day7_tidy %>% 
        filter(outside == bag_desc)
    
    sum(
        d$weight * 
            (map_dbl(d$inside, count_bags) + 1)
    )
}

sprintf(
    "Part 2 answer: the shiny gold bag must contain %s other bags", 
    count_bags("shiny gold")
)
```

# Day 8

## Part 1

Parsing a boot code. Similar to the intcode thing from last year. This looks like a job for a `while` loop with some bookkeeping. Wrap it in a function as it's likely we'll need it again. 

```{r parse_boot_code}
parse_boot_code
```


```{r day8-part1}
day8_input <- tibble(x = read_lines(here::here("data/day8.txt"))) %>% 
    separate(x, into = c("cmd", "value"), sep = " ", convert = TRUE)

parse_boot_code(day8_input) %>% 
    pluck("accumulator") %>% 
    {
        sprintf("Part 1 answer: the accumulator is %s", .)
    }
```

## Part 2

Now we need to fix the corrupted boot code. In other words: for each of the `jmp` and `nop` commands in the boot code we need to figure out if switching it will cause the programme to terminate because it's gone past the end of the code (rather than because it's looped round to something it already visited). 

```{r day8-part2}
to_test <- day8_input %>% 
    rowid_to_column() %>% 
    filter(cmd %in% c("jmp", "nop")) %>% 
    mutate(
        new_cmd = map_chr(cmd, ~ switch(.x, jmp = "nop", nop = "jmp"))
    )

to_test %>% 
    mutate(
        res = map2(
            rowid, 
            new_cmd, 
            function(rowid, new_cmd) {
                tmp <- day8_input
                tmp[rowid, "cmd"] <- new_cmd
                parse_boot_code(tmp)
            }
        )
    ) %>% 
    unnest_wider(res) %>% 
    filter(i > nrow(day8_input)) %>% 
    pull(accumulator) %>% 
    {
        sprintf("Part 2 answer: the accumulator is %s", .)
    }
```

# Day 9

## Part 1

Finding all combinations of the previous 25 values is nice and quick with matrices. 

```{r day9-part1}
day9_input <- read_lines(here::here("data/day9.txt")) %>% 
    as.double()

preamble <- 25L
found_error <- FALSE
i <- preamble + 1L
while(isFALSE(found_error)) {
    prev <- day9_input[seq(i - preamble, i - 1L)]
    summed <- outer(prev, prev, "+")
    if (day9_input[i] %in% summed) {
        i <- i + 1L
    } else {
        found_error <- TRUE
    }
}
day9_part1_ans <- day9_input[i]
sprintf("Part 1 answer is %s", day9_part1_ans)
```

## Part 2

We get told that the range is _contiguous_, so we can loop over the vector just once and keep a running total. If it's too small just raise the upper bound for the range; if it's too large raise the lower bound. 

```{r day9-part2}
lower <- 1
upper <- 1
total <- sum(day9_input[seq(from = lower, to = upper)])
while (total != day9_part1_ans) {
    if (total < day9_part1_ans) {
        upper <- upper + 1
    } else {
        lower <- lower + 1
    }
    total <- sum(day9_input[seq(from = lower, to = upper)])
}
sprintf(
    "Part 2 answer is %s", 
    sum(range(day9_input[lower:upper]))
)
```

# Day 10

## Part 1

This first part is just about counting how many of the jolts have a `diff()` of 1 or 3. Can quickly check the table of all the diffs: 

```{r day10-diff-table}
table(diff(day10_input))
```

There are only diffs of 1 and 3, which simplifies things. 

```{r day10-part1}
day10_input <- read_lines(here::here("data/day10.txt")) %>% 
    as.integer() %>% 
    {
        c(., 0L, max(.) + 3L)
    } %>% 
    sort()

sprintf("Part 1 answer is %s", prod(table(diff(day10_input))))
```

## Part 2

Counting all of the combinations is a non-starter. We know from Part 1 though that there are only diffs of 1 and 3 in the data. And what is the longest run of 1's in the data?

```{r day10-runs}
runs <- rle(diff(day10_input))
max(runs$lengths[runs$values == 1])
```

So then we just multiply each run of ones by the number of ways to rearrange them to make up the length of the run. So if there's a length 3 run of 1's you could rearrange them in 4 ways. There's 1 way to make 1 from 1's; 2 ways to make 2 (1 + 1 or 2); 4 ways to make 3; and 7 ways to make 7. 

```{r day10-part2}
runs$lengths[runs$values == 1] %>% 
    as.integer() %>% 
    {
        c(1, 2, 4, 7)[.]
    } %>% 
    prod() %>% 
    as.character() %>% 
    {
        sprintf("Part 2 answer is %s", .)
    }
```


